Towards an Ecological Trait-data Standard

### 2018-05-22 manuscript/manuscript.Rmd
finalize manuscript v0.8; complete glossary

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2] \footnotemark[3]
- name: Malte Jochum \footnotemark[4]
- name: Gaetane Le Provost \footnotemark[3]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: David Fichtmüller \footnotemark[5]
- name: Anton Güntsch \footnotemark[5]
- name: Pete Manning \footnotemark[3]
- name: Birgitta König-Ries\footnotemark[6]
- name: Nadja K. Simons \footnotemark[7]
csl: amnat.csl
fontsize: 12pt
geometry: margin=1.3in
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

<!-- 3. trait databases have been developed, with standard structure and terms --> 
To this end, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. 
These databases are then published as static data tables (i.e. product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g. TRY, www.try-db.org). Furthermore, these portals map heterogeneous data into a common scheme, and importantly, also offer access control and data usage policies. As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers. 
Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed. These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data are beeing developed.
<!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->

<!-- 4. distributed trait data must be integrated --> 
While these community efforts for centralisation and harmonization are great advancements in biodiversity informatics and data-management, at the same time the number of decentralised trait-datasets is increasing rapidly. This is due to the fact that open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15]., which supports. As a result, an increasing number of individual research projects comply with open-data standards and publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material. <!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
It is likely that data will become increasingly available, but a lack of data standardisation will hamper the efficient re-use and synthesis of published datasets.

In this paper, we review existing trait databases and online portals, as well as initiatives for standardisation. We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users. Based on these considerations, we propose a minimal structure and vocabulary for describing trait-datasets, that is compatible with existing terminology standards for biodiversity data. Finally, we present an R package that assists the harmonisation of trait data from distributed sources.

With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of trait-data standardisation and sketch a roadmap towards an accessible ecological trait data standard.

@@

-->
\scriptsize

@@

Concept | An idea, notion or object that is made explicit in a communication context by *name*, definition, *URI* or other reference (https://www.w3.org/TR/skos-reference/)
Controlled vocabulary | 
Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g. trait data), that is recognized by a large proportion of members of the application context
Terminology | The  body of *terms* used with a particular application in a subject of study, usually formalized in a *thesaurus* or *ontology*
Thesaurus | Controlled vocabulary that provides key *terms* with their associated *concepts* for a specific field or domain of interest [@laporte12] 
Ontology | Controlled vocabulary that (opposed to a *thesaurus*) relates *concepts* to each other by cross-references, e.g. defines a hierarchy of terms; thus a formal model of a the objects and their relationships in a domain of interest [@gruber95]
Semantic web | The concept of connecting data on the internet via well-defined standards and exchange protocols [@berners-lee02]; the World Wide Web Consortium (W3C) defines standards, i.e. specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
Dataset | a set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being harmonized across all data entries;  

@@

Database | a suite of data compiled from multiple *datasets*, i.e. from multiple study contexts or observation types; may take the form of a two-dimensional data table, but mostly is organised in into *relational databases* using database software tools; 

@@

Online database | A relational database that is made accessible on the internet offering forms for filtering and downloading subsets of the data; often using technologies like SQL; some online databases offer access via a *webservice* and an *API* that can be addressed computationally 
Identifier (ID) | A unique label that relates entries within and across datasets; is used to link data tables into a relational database; can be user-specific or as a *URI* be part of a globally valid ontology or thesaurus  
Uniform Resource Identifier (URI) |
Web service | An exchange protocoll to access online databases directly and programmatically, i.e. by calls from a software tool; 
Application Programming Interface (API) | 

@@

Repository | Archive for data files; online repositories make data available for public access, provide metadata and (not always) facilitate citations via DOIs (Digital Object Identifiers)
Metadata | Data documentation representing the higher level information or instructions that describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97]
Darwin Core Standard (Dwc) | Body  of standards containing a glossary of terms intended to facilitate the sharing of information about biological diversity by providing reference definitions, examples, and commentaries. (From http://rs.tdwg.org/dwc/)
Darwin Core Archive (DwC-A) | A file archive (.zip) that contains structured *metadata* and primary data in the form of *two-dimensional data tables* combined into a *relational database* via *identifier* columns. 
Occurrence | 
Measurement | 
Aggregated Value |
Value | 
Fact | 
Method handbook | 
File hosting service | An online platform that hosts *datasets* or entire *repositories* and provides access to a wide audience on the internet; popular examples in biology are Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org;

Table: Glossary of terms from the biodiversity data-management context as they are used in this paper. (draws from @garnier16)

@@

@@

In this section, we review four types of initiatives that are of relevance for trait data standardisation:

@@

In the field of comparative biology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been published in regular journal articles or books. 
With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @gossner15; @ricklefs17].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->

Today, the online publication of such data is greatly facilitated by file hosting services (e.g. Figshare.com). 
Furthermore, these services warrant long-term accessibility, offer DOIs for referencing, and support Public Domain dedication or Creative Commons licenses.
However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk.

Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds for *metadata* documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management. 
As a result, datasets greatly vary in structure and terminology.

@@

A global overview of existing trait data for all taxa and trait types is difficult to obtain. Therefore, in an attempt to collate a list of existing, distributed and unstandardised datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for a current excerpt of this list). We invite data owners and users to add further trait datasets to this spreadsheet.

As it stands, the decentralisation and the lack of data standardisation of low-threshold online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be re-used and combined into broad synthesis analysis.

@@

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with data standardisation. 
With the digitization efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowdsourcing (e.g. www.markmybird.org), data on body measurements are likely to grow exponentially in the near future. 
For example, the VertNet database compiled and harmonized large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources are integrated.

@@

To conclude, there is already a suite of globally available thesauri and ontologies for traits that emerged from standardisation efforts of methodologies and community consensus processes. Albeit definitions in some domains are better covered than others. Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards.
Most importantly, while these defined vocabularies are widely used in biodiversity data management, distributed data repositories of smaller project contexts hardly make use of them. 
A more widespread implementation of ontologies would advance the possibilities to aggregate datasets into databases and reduce noise and uncertainty. 
To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
For example the accessibility of ontologies will increase if advanced webservices, and particularly open Application Programming Interfaces (APIs), are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. 
Software then can assist researchers in linking own data to globally defined concepts.

@@

We suggest that any trait dataset that is published on online repositories should draw its column names and field entries from the defined vocabulary where possible. 
The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17] each entry describes a trait value (i.e. quality) observed on an an individual or population (i.e. entity), of a biological taxon. 
It is implicit to the use of the vocabulary to use a two-dimensional observation long-table format for the data, rather than a species$\times$trait matrix. As the long-table format draws from a defined set of columns, the merging of datasets is  easier. Long-table datasets also purport multiple advantages for data manipulation [e.g. filtering, sub-setting and aggregating data, @wickham14; Fig. 2].

Well-defined identifiers ('IDs') are key elements to structure the datasets and relate them to complementing datasets, if necessary.
For instance, for occurrence level data where multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link several measurements across the rows of the dataset. 
Similarly, multivariate measurements, for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
In literature data, summarized traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
In larger compilations, a 'datasetID' allow to trace data origin to the primary source. 
Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via Uniform Resource Identifiers (URIs), which point to external terminology services (see above for resources). 
Two-dimensional spreadsheets are however limited in the number and complexity of co-variates they can contain. As such, for datasets containing multi-layered information on observations, traits, taxa and environmental context, the use of relational datatabase structures may be indicated, like the generic trait database structure proposed by @kattge11a or the TraitBank structure proposed by @parr16. 
The trade-off is user-side readability and handling in a single table *vs.* avoidance of content duplication and redundancy in a relational database. 
The standard vocabulary proposed here may still be applied to describe columns within the individual data tables of relational databases.

@@

To qualify as trait data according to the definition provided above, where each row is the reported measurement or fact for a single observation, the following columns are required at minimum: 
1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
2. a descriptive trait name (`traitName`) that links to a well-defined definition; 
3. the scientific taxon for which the measurement or fact was obtained (`scientificName`). 
For these core values, unambiguous and self-explanatory vocabularies for trait names and taxa are recommended. 
However, to ensure compatibility with existing databases or analytical code, it might be necessary to use abbreviations or user-specific identifiers for `scientificName`  and `traitName` instead. 
<!--### Relate to global taxon and trait definitions -->
In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up table of traits. 
This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxon (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively). 
For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'.  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11]. This redundancy of data allows for continuity for data owners while also ensuring quality checks and comparability for the data user.

@@

Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event. 
The data standard provides three extensions of the vocabulary that should be used to describe this information:

@@

Wherever possible, the colum `traitID` should point to a public trait definition, ideally as part of an unambiguous ontology according to semantic-web standards. 
If no globally available trait definition exists as a reference, trait-datasets should always be accompanied by a dataset-specific glossary or thesaurus of traits. Such a 'controlled vocabulary' would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format  of measured values or reported facts (e.g. units or legit factor levels). 
By providing a minimal vocabulary for trait thesauri (see https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the publication of trait thesauri developed for a specific project context, which can be referenced in the trait dataset. The thesaurus might accompany the core data in a separate file (e.g. within a Darwin Core Archive), or be published on any other stable internet address (ideally providing stable URIs). 
<!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
See our online reference for recommendations how to publish a minimal thesaurus compliant with semantic web standards (https://ecologicaltraitdata.github.io/ETS/thesauri.html). We also provide an example thesaurus on Github (http://ecologicaltraitdata.github.io/TraitDataList/), which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project [@fischer10].

@@

The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement longtable format (Fig. 3). This function also maps column names into terms provided in the trait data standard and adds metadata as attributes to the output object. This example converts an own file 'data.csv' into a dataset of long-table structure that employs the standard vocabulary for core data:

@@

The parameter 'traits' lists column names that contain trait values. The column containing taxon names is given in parameter 'taxa'. Note that the parameter 'keep' specifies and renames any data that should be maintained in the output. The parameter 'units' is used to specify the input units of measurement. 
In order to map user-provided names to unambiguous and globally unique identifiers, the function `standardize.taxonomy()` matches scientific taxon names automatically to the GBIF Backbone Taxonomy and adds the column `taxonID` to the core data. 
<!-- BKR: In the example below, this looks like a manual matching. What is done automatically here and which information does the user have to provide?  FDS: taxon is matched automatically, traits not. lack of trait APIs and webservices. --> 
<!-- Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e. a thesaurus of traits. The same function harmonises trait values by converting values into target units and unified factor levels. 
<!-- BKR: Have you thought about linking to unit ontologies here?
AO: This paper is a good reference and gives an idea of which ontology could be used;
http://www.semantic-web-journal.net/content/comparison-and-evaluation-ontologies-units-measurement
FDS: might be referred to in the section about thesauri -->

The following example harmonizes traits based on a minimal thesaurus, referencing trait names with globally valid URIs defined by the BETSI thesaurus of soil invertebrate traits:

@@

The function `as.thesaurus()` provides a structured object that is required by the function `standardize.traits()`. Other ways of defining a thesaurus object are documented in the package vignette and function documentation (`?as.thesaurus`). Future iterations of the R package will aim at automatizing the generation of trait thesaurus objects from globally available Ontologies. 
The package functions form a toolchain where each function can be piped as an input into the next. A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing trait data as simple as:

@@

Any future development of the Vocabulary and the R-package is coordinated via https://github.com/EcologicalTraitData/.

@@

The living spreadsheet on existing trait datasets and databases is found at https://goo.gl/QxzfHy.

>>>>

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2] \footnotemark[3]
- name: Malte Jochum \footnotemark[4]
- name: Gaetane Le Provost \footnotemark[3]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: David Fichtmüller \footnotemark[5]
- name: Anton Güntsch \footnotemark[5]
- name: Pete Manning \footnotemark[3]
- name: Birgitta König-Ries\footnotemark[6]
- name: Nadja K. Simons \footnotemark[7]
csl: amnat.csl
fontsize: 11pt
geometry: margin=1.3in
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

<!-- 3. trait databases have been developed, with standard structure and terms --> 
To this end, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. 
These databases are then published as static data tables (i.e. product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g. TRY, www.try-db.org). Furthermore, these portals map heterogeneous data into a common scheme, and importantly, also offer access control and data usage policies. As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers. 
Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed. These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data are beeing developed.
<!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
<!-- 4. distributed trait data must be integrated --> 
While these community efforts for centralisation and harmonization are great advancements in biodiversity informatics and data-management, at the same time the number of decentralised trait-datasets is increasing rapidly. This is due to the fact that open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12; @swan12; @allison15]. As a result, an increasing number of individual research projects comply with open-data standards and publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material. <!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
It is likely that data will become increasingly available, but a lack of data standardisation will hamper the efficient re-use and synthesis of published datasets.

In this paper, we review existing trait databases and online portals, as well as initiatives for standardisation. We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users. Based on these considerations, we propose a minimal structure and vocabulary for describing trait-datasets, that is compatible with existing terminology standards for biodiversity data. Finally, we present an R package that assists the harmonisation of trait data from distributed sources. 
With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of trait-data standardisation and sketch a roadmap towards an accessible ecological trait data standard.

@@

-->
\newpage

\tiny

@@

Concept | An idea, notion or object that is made explicit in an information context by *name*, definition, *URI* or other reference (https://www.w3.org/TR/skos-reference/)
Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a praticular context, while no unlisted entries are accepted
Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalized in a *thesaurus* or *ontology*
Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g. trait data), that is recognised by a large proportion of members of the application context
Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* for a specific field or domain of interest [@laporte13] 
Ontology | *Controlled vocabulary* that (opposed to a *thesaurus*) relates *concepts* to each other by cross-references, e.g. defines a hierarchy of terms; thus a formal model of a the objects and their relationships in a domain of interest [@gruber95]
Semantic web | The concept of connecting data on the internet via well-defined *data standards* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e. specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
Dataset | a set of measurements and observations; often originating from a single experimental set-up or study context; can be considered as being homogeneous across all data entries

@@

Database | a suite of data compiled from multiple *datasets*, i.e. from multiple study contexts or observation types; may take the form of a two-dimensional data table, but mostly is organised in into *relational databases* using database software; 

@@

Online database | A *relational database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; often using technologies like SQL; some online databases offer access via a *webservice* and an *API* that can be addressed computationally 
Identifier (ID) | A unique label that relates entries within and across *datasets*; is used to connect *data tables* into a *relational database*; can be user-specific or, as a *URI*, point to a globally valid *ontology* or *thesaurus*  
Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; an example is 'http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length'
Webservice | An exchange protocoll to access *online databases* directly and programmatically, i.e. by calls from a software tool
Application Programming Interface (API) |  a set of clearly defined methods of communication between  software components, e.g. client software and the *webservice* of an online databases; APIs are usually documented on the website of a database provider

@@

File repository | a structure for data files, for archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via DOIs (Digital Object Identifiers) 
File-hosting service | An online platform that hosts *datasets* or entire *repositories* and provides access to a wide audience on the internet; examples in biology are Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org
Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97]
Darwin Core Standard (Dwc) | Body  of *terminologies* providing terms intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/)
Darwin Core Archive (DwC-A) | A file archive, or *repository* structure, that contains *metadata* and primary data combined into a *relational database* via *identifier* columns. 
Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures. 
Occurrence | A single observation instance of a taxon, i.e. an organism at a particular place at a particular time (http://rs.tdwg.org/dwc/terms/Organism)

Table: Glossary of terms from the biodiversity data-management context as they are used in this paper. (draws from @garnier17)

<!--
Measurement |  
Aggregated Value |
Value | 
Fact | 
-->

@@

\newpage

@@

In this section, we review four types of initiatives that are of relevance for trait data standardisation (see Glossary in Table 1 for italicised terms):

@@

In the field of comparative biology, morphological traits such as plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values  have been measured for decades, and have been published in regular journal articles or books. 
With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @gossner15; @ricklefs17].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
Today, the online publication of such data is greatly facilitated by *file hosting services* (e.g. Figshare.com). 
Furthermore, these services warrant long-term accessibility, offer DOIs for referencing, and support Public Domain dedication or Creative Commons licenses.
However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk. 
Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds for *metadata* documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management. 
As a result, datasets greatly vary in structure and terminology.

@@

A global overview of existing trait data for all taxa and trait types is difficult to obtain. Therefore, in an attempt to collate a list of existing, distributed and unstandardised datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for a current excerpt of this list). We invite data owners and users to add further trait datasets to this spreadsheet.  
As it stands, the decentralisation and the lack of data standardisation of low-threshold online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be re-used and combined into broad synthesis analysis.

@@

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with data standardisation. 
With the digitization efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12a], supported by citizen science crowdsourcing (e.g. www.markmybird.org), data on body measurements are likely to grow exponentially in the near future. 
For example, the VertNet database compiled and harmonized large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources are integrated.

@@

To conclude, there is already a suite of globally available thesauri and ontologies for traits that emerged from standardisation efforts of methodologies and community consensus processes. Albeit definitions in some domains are better covered than others. Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards.
Most importantly, while these defined vocabularies are widely used in biodiversity data management, distributed data repositories of smaller project contexts hardly make use of them. 
A more widespread implementation of ontologies would advance the possibilities to aggregate datasets into databases and reduce noise and uncertainty. 
To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
For example the accessibility of ontologies will increase if advanced webservices, and particularly open *Application Programming Interfaces (APIs)*, are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. 
Software then can assist researchers in linking own data to globally defined concepts.

@@

We suggest that any trait dataset that is published on online repositories should draw its column names and field entries from the defined vocabulary where possible.
The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17] each entry describes a trait value (i.e. quality) observed on an an individual or population (i.e. entity), of a biological taxon.
It is implicit to the use of the vocabulary to use a two-dimensional observation long-table format for the data (Fig. 2 *b*), rather than a species$\times$trait matrix (Fig. 2 *a*). As the long-table format draws from a defined set of columns, merging datasets is easier. Long-table datasets also purport multiple advantages for data manipulation [e.g. filtering, sub-setting and aggregating data, @wickham14].

Well-defined identifiers ('IDs') are key elements to structure the datasets and relate them to complementing datasets, if necessary (Fig. 2 *c* & *d*).
For instance, for occurrence level data where multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link several measurements across the rows of the dataset. 
Similarly, multivariate measurements, for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
In literature data, summarized traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
In larger compilations, a 'datasetID' allow to trace data origin to the primary source. 
Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via Uniform Resource Identifiers (URIs), which point to external terminology services (see above for resources). 
Two-dimensional spreadsheets are however limited in the number and complexity of co-variates they can contain. As such, for datasets containing multi-layered information on observations, traits, taxa and environmental context, the use of relational datatabase structures may be indicated, like the generic trait database structure proposed by @kattge11a or the TraitBank structure proposed by @parr16. 
The trade-off is user-side readability and handling in a single table *vs.* avoidance of content duplication and redundancy in a relational database. 
The standard vocabulary proposed here may still be applied to describe columns within the individual data tables of relational databases.

@@

To qualify as trait data according to the definition provided above, where each row is the reported measurement or fact for a single observation, the following columns are required at minimum  (Fig. 2 *b*):
1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
2. a descriptive trait name (`traitName`) that links to a well-defined definition; 
3. the scientific taxon for which the measurement or fact was obtained (`scientificName`). 
For these core values, unambiguous and self-explanatory vocabularies for trait names and taxa are recommended. 
However, to ensure compatibility with existing databases or analytical code, it might be necessary to use abbreviations or user-specific identifiers for `scientificName`  and `traitName` instead. 
<!--### Relate to global taxon and trait definitions -->
In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up table of traits. 
This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxon (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively). 
For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'.  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11]. This redundancy of data allows for continuity for data owners while also ensuring quality checks and comparability for the data user.

@@

Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event. 
The data standard provides three extensions of the vocabulary that should be used to describe this information (Fig. 2 *d*):

@@

Wherever possible, the column `traitID` should point to a public trait definition, ideally as part of an unambiguous ontology according to semantic-web standards. 
If no globally available trait definition exists as a reference, trait-datasets should always be accompanied by a dataset-specific glossary or thesaurus of traits. Such a 'controlled vocabulary' would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format  of measured values or reported facts (e.g. units or legit factor levels). 
By providing a minimal vocabulary for trait thesauri (see https://ecologicaltraitdata.github.io/ETS/#terms-for-thesauri), we hope to facilitate the publication of trait thesauri developed for a specific project context, which can be referenced in the trait dataset. The thesaurus might accompany the core data in a separate file (e.g. within a Darwin Core Archive), or be published on any other stable internet address (ideally providing stable URIs). 
<!-- PM: Is this possible in practice?- in plant traits slightly different methods makes some traits comparable conceptually but not in practice, and there are lots of very minor variations of similar traits- see the vast list of traits in TRY for example -->
See our online reference for recommendations how to publish a minimal thesaurus compliant with semantic web standards (https://ecologicaltraitdata.github.io/ETS/thesauri.html). We also provide an example thesaurus on Github (http://ecologicaltraitdata.github.io/TraitDataList/), which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project [@fischer10].

@@

The key function of the package is `as.traitdata()` which moulds a species-trait-matrix or occurrence table into a measurement long-table format (Fig. 3). This function also maps column names into terms provided in the trait data standard and adds metadata as attributes to the output object. This example converts an own file 'data.csv' into a dataset of long-table structure that employs the standard vocabulary for core data:

@@

The parameter 'traits' lists column names that contain trait values. The column containing taxon names is given in parameter 'taxa'. Note that the parameter 'keep' specifies and renames any data that should be maintained in the output. The parameter 'units' is used to specify the input units of measurement. 
In order to map user-provided names to unambiguous and globally unique identifiers, the function `standardize.taxonomy()` matches scientific taxon names automatically to the GBIF Backbone Taxonomy and adds the column `taxonID` to the core data  (Fig. 3). 
<!-- BKR: In the example below, this looks like a manual matching. What is done automatically here and which information does the user have to provide?  FDS: taxon is matched automatically, traits not. lack of trait APIs and webservices. --> 
<!-- Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e. a thesaurus of traits. The same function harmonises trait values by converting values into target units and unified factor levels. 
<!-- BKR: Have you thought about linking to unit ontologies here?
AO: This paper is a good reference and gives an idea of which ontology could be used;
http://www.semantic-web-journal.net/content/comparison-and-evaluation-ontologies-units-measurement
FDS: might be referred to in the section about thesauri -->

The following example harmonises traits based on a minimal thesaurus, referencing trait names with globally valid URIs defined by the BETSI thesaurus of soil invertebrate traits:

@@

The function `as.thesaurus()` provides a structured object that is required by the function `standardize.traits()`  (Fig. 3). Other ways of defining a thesaurus object are documented in the package vignette and function documentation (`?as.thesaurus`). Future iterations of the R package will aim at automatizing the generation of trait thesaurus objects from globally available Ontologies. 
The package functions form a tool-chain where each function can be piped as an input into the next. A wrapper function `standardize()` applies all functions sequentially, making transferring and harmonizing trait data as simple as:

@@

Any future development of the vocabulary and the R-package is coordinated via https://github.com/EcologicalTraitData/.

@@

An example of a published trait thesaurus for the Biodiversity Exploratories (www.biodiversity-exploratories.de) can be found at https://ecologicaltraitdata.github.io/TraitDataList/.

The living spreadsheet on existing trait datasets and databases can be found at https://goo.gl/QxzfHy.
